Many claims of group polarization are plausibly false due to inappropriate statistics

### 2024-01-18 GroupPolarizationStatmod/paper/ms.tex
updated abstract

\begin{abstract}
  \noindent
  % Psychology is beset by three crises: the reproducibility crisis, the
  % generalizability crisis, and the theory crisis. New verbal theories for various
  % phenomena are constantly being developed orthogonal to other theories, all based on
  % mostly unreproducible analyses; even if a positive result is in some sense ``true''
  % it may not account for all the variance present in a set of observations---studies
  % thus fail to generalize because their results do not account for this variance.
  % There is another related problem: a phenomenon may be ``reproducible'' only
  % because studies of the phenomenon systematically misuse statistics in such a way
  % that induces false positives for each succeeding ``replication.'' 
  \emph{Group polarization} refers to the observation that like-minded novel groups tend to
  become more extreme in their opinions after discussion as group members found consensus.
  \emph{Consensus} here refers to decreasing opinion variance; we further define
  \emph{simple consensus} as a process where group opinion variance decreases but
  the mean opinion does not change.
  Here we show that many published experimental detections of group polarization are explained
  equally rigorously as \emph{simple consensus}. 
  The reports we analyze assumed in their statistical analyses 
  that both latent psychological opinions \emph{and} the ordinal 
  reported opinions (e.g., on a Likert-type scale) are continuous. We use
  a generative model to demonstrate that 95\% of sixty experimental conditions
  across ten foundational group polarization papers report plausibly false detections of
  group polarization. We show that simulated data from these plausibly false
  cases often fools a $t$-test into reporting significant group polarizaiton
  effects, but a Bayesian implementation of the more appropriate ordered-probit
  model reliably identifies these data as
  representing simple consensus, not group polarization.
  Our results demonstrate the need for a new wave of group polarization research
  that uses rigorous statistical methods appropriate for the study design. 
  Our approach can be used beyond group polarization
  research to analyze results from any experimental paradigm that may use statistical models that
  are poorly matched to the study design. Replications are worthless if the statistics are wrong.

>>>>

\begin{abstract}
  \noindent
  % Psychology is beset by three crises: the reproducibility crisis, the
  % generalizability crisis, and the theory crisis. New verbal theories for various
  % phenomena are constantly being developed orthogonal to other theories, all based on
  % mostly unreproducible analyses; even if a positive result is in some sense ``true''
  % it may not account for all the variance present in a set of observations---studies
  % thus fail to generalize because their results do not account for this variance.
  % There is another related problem: a phenomenon may be ``reproducible'' only
  % because studies of the phenomenon systematically misuse statistics in such a way
  % that induces false positives for each succeeding ``replication.'' 
  \emph{Group polarization} is a theoretical process where like-minded novel group members 
  become more extreme in their opinions after group discussion, but also find \emph{consensus}, reducing 
  group-level opinion variance.
  \emph{Simple consensus} may occur instead, where opinion variance decreases but
  the mean group opinion does not change.
  Here we show that many published detections of group polarization are explained
  equally rigorously as simple consensus.
  This is the case because experimenters failed
  to account for the measurement of continous-valued psychological opinions 
  as ordinal-valued reported opinions, e.g., on a Likert-type scale. 
  This creates ceiling effects that mask whether extreme opinions become less extreme.
  We developed a generative model to demonstrate that 95\% of sixty experimental conditions
  across ten group polarization papers report plausibly false detections of
  group polarization. Our approach can be used to evaluate whether other detections of
  group polarization may be plausibly false---our approach could be applied beyond
  group polarization as well. This work also provides a
  framework for new group polarization studies that could
  share a common, rigorous, and interoperable study design to multiply our knowledge
  of this potentially important group process, avoiding the historically fragmented
  approach using disparate, incommensurate study designs.
